Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-19 concurrency/CONCUR 2023/paper.tex
write more about proving insert of give-up template

\subsubsection{Lock-coupling Template}
The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal. Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.

@@

\subsubsection{Give-Up Template}
The give-up template is an algorithm for implementing concurrent search in a tree data structure. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template takes an \emph{optimistic concurrency control} approach \lb{c.f. RCU}, only acquiring locks when loading or storing to a node. This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios. Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path. If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again. This happens when another concurrent thread restructures the data structure, invalidating the state of the current thread in between releasing one lock and acquiring another. For instance, the node we are traveling to may be moved to another part of the data structure intended to hold a different set of keys, or even deleted, before we acquire its lock.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			 \forall \  \texttt{m}. \left\langle
				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
			\right\rangle
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}
%\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$. But now we can ignore the information of $\texttt{my\_half}$.  }

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  \texttt{m}. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{m}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
				\\ 
				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

We have defined the loop invariant for the given template of the \texttt{traverse} function as follows: 
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{align*}

@@

In the \texttt{then} branch, when the \texttt{traverse} fails to find the next node, resulting in $\texttt{res} = \texttt{false}$, it implies that $\texttt{pn->n'->t} \neq \texttt{NULL}$. This indicates that the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}); this acts as the linearization point for the insertion. Conversely, when the \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, it implies $\texttt{pn->n'->t} = \texttt{NULL}$ (line 9 in Figure \ref{insert_lock}). We then utilize the \texttt{insertOp} specification, and with enough information, we can conclude that inserting the key at this location will alter the abstract state from an unknown $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$. In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair.

@@

>>>>

\subsubsection{Lock-coupling Template}
\label{lock-coupling-algo}
The lock-coupling template employs the hand-over-hand locking scheme to prevent interference from other threads during traversal. Each thread continuously maintains at least one lock during traversal between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search.

@@

\subsubsection{Give-Up Template}
\label{give-up-algo}
The give-up template is an algorithm for implementing concurrent search in a tree data structure. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template takes an \emph{optimistic concurrency control} approach \lb{c.f. RCU}, only acquiring locks when loading or storing to a node. This reduces the amount of time that locks are held and makes the algorithm more scalable in highly-contended scenarios. Each node has a range field that serves as a reference point for threads to determine if they are still on the correct path. If a thread finds that it has deviated from the expected path, it abandons the node by releasing the lock and returns to the root node to start again. This happens when another concurrent thread restructures the data structure, invalidating the state of the current thread in between releasing one lock and acquiring another. For instance, the node we are traveling to may be moved to another part of the data structure intended to hold a different set of keys, or even deleted, before we acquire its lock.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			 \forall \  \texttt{m}. \left\langle
				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}
%\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$. But now we can ignore the information of $\texttt{my\_half}$.  }

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  \texttt{m}. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{m}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
				\\ 
				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

In contrast to the lock-coupling style, where each lock is associated with a lock-invariant—an arbitrary separation logic predicate representing the protected resource acquired by a thread when the lock is obtained and must be restored upon lock release—the give-up template has a different approach. In this method, every node possesses a lock bit at location $\ell$, which is set to \texttt{True} if and only if the node is locked. This protocol is encoded into the definition of the abstract state $\treerep$. The lock protects resources denoted by \texttt{R}, which can be removed from $\treerep$ by threads when the node is locked.

Before proving the \texttt{traverse} function, let's define the loop invariant for the given template of the \texttt{traverse} function as follows:
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{align*}

@@

In the \texttt{then} branch, when the \texttt{traverse} fails to find the next node, resulting in $\texttt{res} = \texttt{false}$, it implies that $\texttt{pn->n'->t} \neq \texttt{NULL}$. This indicates that the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}); this acts as the linearization point for the insertion. Conversely, when the \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, it implies $\texttt{pn->n'->t} = \texttt{NULL}$ (line 9 in Figure \ref{insert_lock}). We then utilize the \texttt{insertOp} specification, and with enough information, we can conclude that inserting the key at this location will alter the abstract state from an unknown $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$. In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair. 
Next, we release (line 11) the node \texttt{n} and return it to the shared state. Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.

@@

Using the give-up template, the \texttt{insert} method in Figure \ref{insert_giveup} is not significantly different from the \texttt{insert} method of the lock-coupling template.  The primary distinction is that it does not require acquiring a lock before calling the \texttt{traverse} function. Furthermore, it utilizes an auxiliary \texttt{traverse} function, as illustrated in Figure \ref{fig:traverse_giveup}.  This function shares the same objective as the previously discussed \texttt{traverse} function in the lock-coupling approach.

The give-up template retains the specification of the \texttt{findNext} helper function from the lock-coupling template while introducing an extra helper function called \texttt{inRange}. Additionally, the \texttt{insertOp} specification exhibits subtle differences from that of the lock-coupling template, as it incorporates two new fields,  \texttt{min} and \texttt{max},  into each \texttt{node\_t} type node, as showcased in Section  \ref{give-up-algo}.

The \texttt{inRange} specification (Figure \ref{fig:inRange_giveup}) simply necessitates that \texttt{inRange(pn, k)}  yields a boolean value, \texttt{res}. When \texttt{res} is \texttt{True}, it signifies that \texttt{k} is within the range of \texttt{(min, max)} associated with the node \texttt{n}. 
Conversely, \texttt{res} is \texttt{False}, it indicates that \texttt{k} is outside the boundaries of \texttt{(min, max)}.

As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) exhibits slight differences from that of the lock-coupling template, primarily due to the incorporation of two extra fields. As a result, node \texttt{n} needs point-to predicates for these new fields. Moreover, the current node's \texttt{l} and \texttt{r} point to two new empty child nodes with associated locks and ranges: \texttt{(NULL, lk1, (min, k))} and \texttt{(NULL, lk2, (k, max))}.  In the context of binary search trees, a node inherits its range from its parent based on the parent's key; thus, if node \texttt{n} has range \texttt{(min, max)} and key \texttt{k}, its left child will have the range \texttt{(min, k)} and its right child the range \texttt{(k, max)}.
